\textcolor{sage}{Does Hospital Leadership Matter? \newline Evidence from Pay-for-Performance}

### 2025-11-10 Paper/paperdraft_fall2025.tex

    \section{Theoretical Framework}\label{sec:theory}

    The first prediction of the model is that, all else held equal, hospitals that value profit will have a lower quality than those who value patient welfare in a fee-for-service setting. Second, a hospital's change in quality once pay-for-performance is enacted depends on how much weight the hospital places on profit vs. patient welfare. Particularly, hospitals with higher weight on profit respond more than hospitals with higher weight on patient welfare. Therefore, factors that affect weight $\alpha$ will be correlated with choice of quality, and with how quality changes under new pay-for-performance incentives. In this paper, I test the hypothesis that clinical experience on a hospital's leadership team is a factor that is correlated with weight $\alpha$, and thus affects response to pay-for-performance incentives.

@@

    The specific aspects of leadership that link executives to employees to patient outcomes are nearly impossible to observe. As the data on this topic progresses, these granular details of leadership will become more accessible to analyze. For the purpose of this paper, it is important for policymakers to understand why there is a behavioral difference in clinician versus non-clinician leaders, as this might affect the types of incentives they will respond to.

    There are several hypotheses that I test given the data available. First, as the theoretical predictions in Section \ref{sec:theory} state, it is possible that clinicians are less financially driven than non-clinicians, and thus respond less drastically to financial incentives. I test this empirically in Section \ref{sec:forprofit}, comparing the response of nonprofits with different leadership teams to the response of for-profit hospitals on average. Second, I investigate whether clinicians are actively managing hospitals differently or whether they are a signal of underlying preferences of the hospital by leveraging the timing of clinical leaders relative to the incentives in Section \ref{sec:sig_man}. Third, in Section \ref{sec:selection}, I consider whether different types of executives might have different propensities to engage in "gaming" through selective patient practices. In Section \ref{sec:conditions}, I dig deeper into the specific conditions driving the main result, and whether these can help uncover the types of leadership changes made. Similarly, I analyze whether different executives change investment behavior in response to the incentives in Section \ref{sec:investment}. Finally, I consider the altruism versus expertise argument using a heterogeneity analysis of different physician specialties in Section \ref{sec:expertise}.

    \subsection{Financial vs. Patient Driven}\label{sec:forprofit}

    The typical theoretical prediction surrounding quality and pay-for-performance given different valuations of finances vs. patient welfare, shown in Section \ref{sec:theory}, certainly support the idea that, if all else is equal, clinician leaders value patients at baseline more than non-clinician leaders. To investigate this hypothesis empirically, I merge all general for-profit hospitals in the AHA survey to my sample of nonprofit hospitals. I then compare readmission and mortality rates of always clinical executives and never clinical executives to for-profit hospitals after the incentive change.

    While I do not have information on leadership teams of for-profit hospitals, on average, they have revealed through their ownership status that they more heavily weigh profit. Thus, comparing different executive team behavior to for-profit behavior helps reveal how profit driven certain leadership teams are. Thus, I estimate the weighted synthetic difference-in-differences version of

    \subsection{Signaling vs. Managing}\label{sec:sig_man}

    Given that having clinically trained executives does affect hospital response to financial incentives on quality, an important distinction is whether the existing clinical leaders \textit{reveal} the underlying objectives of the hospital, or \textit{manage} the hospital differently. Up to this point, I have limited the sample of nonprofit hospitals to those that don't change their propensity to hire a clinically trained executive over the sample period, which combines any signaling and managing effects. However, under the assumption that hiring or firing these types of executives is not correlated with the pay-for-performance policy changes, such changes can be leveraged to decompose the effect into signaling versus managing. Since this decomposition depends heavily on the assumption that clinical executive hiring is not endogenous, I first investigate leadership changes in response to the policy change. I provide details of the decomposition estimation in Section \ref{sec:decomp}

@@

    \subsection{Specific Conditions (rename later)}\label{sec:conditions}

    \subsection{Altruism vs. Expertise}\label{sec:expertise}

    While most doctors take a hippocratic oath to provide care with integrity and compassion, doctors with training in certain specialties may be more equipped to improve care for certain patients. Since the pay-for-performance incentive programs directly target three conditions, physicians with knowledge of these conditions may be especially suited to respond to the policies. Specifically, a pulmonologist typically treats pneumonia, and a cardiologist typically treats heart failure and heart attack patients, which are the three conditions considered in the HRRP penalties. Both of these specialties are under the umbrella of internal medicine, which is the most granular specialty type in my data.

    I limit the sample of hospitals to those who always have some type of clinical executive over the sample period. Then, I estimate the difference in readmission and mortality rates after pay-for-performance among hospitals who do and do not have an internal medicine clinical executive. I lose a significant fraction of the sample with this limitation: there are 60 hospitals that consistently have an internal medicine executive or only a different specialty executive. The summary statistics for these samples are in Appendix \ref{app:sumstats}. I present the results from this comparison in Figure \ref{fig:specialty}.

@@

>>>>

@@

    While uncovering that there is some foundational difference in clinical and non-clinical hospital leaders is important in its own right, it is especially relevant to understand why this difference might come about. In this Section, I present several theoretical predictions consistent with the observed difference in quality after pay-for-performance incentives, and  test the predictions empirically. First, I investigate whether clinicians are actively managing hospitals differently or whether they are a signal of underlying preferences of the hospital by leveraging the timing of clinical leaders relative to the incentives in Section \ref{sec:sig_man}. Second, the difference could be driven by the altruism, or patient-centered focus, of clinicians. I investigate this in Section \ref{sec:forprofit}, comparing the response of nonprofits with different leadership teams to the response of for-profit hospitals on average. Third, I consider the altruism versus expertise argument using a heterogeneity analysis of different physician specialties in Section \ref{sec:expertise}. In Section \ref{sec:selection}, I consider whether different types of executives might have different propensities to engage in "gaming" through selective patient practices. Finally, I analyze whether different executives change investment behavior in response to the incentives in Section \ref{sec:investment}.

    \subsection{Signaling versus Managing}\label{sec:sig_man}

    Given that having clinically trained executives does affect hospital response to financial incentives on quality, an important distinction is whether the existing clinical leaders \textit{reveal} the underlying objectives of the hospital, or \textit{manage} the hospital differently. In the main analysis, I limited the sample of nonprofit hospitals to those that don't change their propensity to hire a clinically trained executive over the sample period, which combines any signaling and managing effects. However, under the assumption that hiring or firing these types of executives is not correlated with the pay-for-performance policy changes, such changes can be leveraged to decompose the effect into signaling versus managing. Specifically, if the difference is driven by underlying preferences of the hospital, the observed difference in quality will still hold whether the hospital had a clinical executive at any point in time, even if they were not a leader in 2012 when the policies took effect. Alternatively, if the observed difference is driven by managing, then the difference should only be observed when there is a clinician in leadership during the policy enactment. Since this decomposition depends heavily on the assumption that clinical executive hiring is not endogenous, I first investigate leadership changes in response to the policy change. I provide details of the decomposition estimation in Section \ref{sec:decomp}

@@

    \subsection{Financially Driven versus Patient Driven}\label{sec:forprofit}

    The first prediction of the model is that, all else held equal, hospitals that value profit will have a lower quality than those who value patient welfare in a fee-for-service setting. Second, a hospital's change in quality once pay-for-performance is enacted depends on how much weight the hospital places on profit vs. patient welfare. Particularly, hospitals with higher weight on profit respond more than hospitals with higher weight on patient welfare. Therefore, factors that affect weight $\alpha$ will be correlated with choice of quality, and with how quality changes under new pay-for-performance incentives. In this paper, I test the hypothesis that clinical experience on a hospital's leadership team is a factor that is correlated with weight $\alpha$, and thus affects response to pay-for-performance incentives. To investigate this hypothesis empirically, I merge all general for-profit hospitals in the AHA survey to my sample of nonprofit hospitals. I then compare readmission and mortality rates of always clinical executives and never clinical executives to for-profit hospitals after the incentive change.

    While I do not have information on leadership teams of for-profit hospitals, on average, they have revealed through their ownership status that they more heavily weigh profit, and therefore have an $\alpha$ close to 1. This is not inconsistent with previous work positing that nonprofits and for-profits act similarly on average (\cite{duggan2002hospital}). In fact, my data is consistent with this idea when nonprofits are not broken down by leadership teams. However, comparing different executive team behavior to for-profit behavior helps reveal how profit driven certain leadership teams are, which could actually help us understand why there are similarities between nonprofits and for-profits on average. I estimate the weighted synthetic difference-in-differences version of

    \subsection{Driven by Expertise}\label{sec:expertise}

    It may be that the observed difference in quality after pay-for-performance could be correlated with the expertise of leadership teams. While most doctors take a hippocratic oath to provide care with integrity and compassion, doctors with training in certain specialties may be more equipped to improve care for certain patients. Since the pay-for-performance incentive programs directly target three conditions, physicians with knowledge of these conditions may be especially suited to respond to the policies. Specifically, a pulmonologist typically treats pneumonia, and a cardiologist typically treats heart failure and heart attack patients, which are the three conditions considered in the HRRP penalties. Both of these specialties are under the umbrella of internal medicine, which is the most granular specialty type in my data.

    Consider an extension to the model discussed in Section \ref{sec:forprofit}, where I now introduce a measure of expertise $\lambda$. To capture the role of hospital leadership expertise in shaping quality decisions, we introduce a binary variable $\lambda \in \{0,1\}$, where $\lambda=1$ indicates high expertise and $\lambda=0$ indicates low expertise. Expertise affects the marginal cost of quality improvements: hospitals with high expertise can implement quality enhancements more efficiently, reducing the effective cost of quality.

    The hospital’s objective function becomes:
    \begin{equation}
        \max_{\theta} \alpha \big[R(\theta) - \frac{c(\theta)}{1+\lambda}\big] + (1 - \alpha) u(\theta),
    \end{equation}

    \noindent where $R(\theta)$ is revenue, which depends on quality,
    $c(\theta)(1 + \lambda)$ is the cost of quality, increasing and convex in $\theta$, scaled by expertise, $u(\theta)$ is patient welfare, increasing and concave in $\theta$, $\alpha \in [0,1]$ is the weight on profit relative to patient welfare. The hospital's first-order condition under fee-for-service and pay-for-performance is then:

    \[
    F_{ffs}(\theta,\alpha,\lambda) = (1-\alpha)u'(\theta) - \alpha\frac{c'(\theta)}{1+\lambda} = 0,
    \]

    \[
    F_{p4p}(\theta,\alpha,\lambda) = (1-\alpha)u'(\theta) - \alpha\Bigg[\frac{c'(\theta)}{1+\lambda} - R'(\theta)\Bigg] = 0.
    \]

    \noindent I make the following assumptions on patient welfare, cost, and revenue: $u'(\theta) > 0,\; u''(\theta) < 0,\; c'(\theta) > 0,\; c''(\theta) > 0,\; R'(\theta) > 0,\; R''(\theta) \approx 0$. By the Implicit Function Theorem.
    \begin{align}
    \frac{d\theta}{d\lambda} &= -\frac{\partial F/\partial \lambda}{\partial F/\partial \theta}\\
    &= -\frac{\alpha c'(\theta)/(1+\lambda)^2}{(1-\alpha)u''(\theta)-\alpha\big[c''(\theta)/(1+\lambda)-R''(\theta)\big]}\\
    &>0. 
    \end{align}

    I now want to compare the change in $\theta$ when going from fee-for-service to pay-for-performance depending on different levels of $\lambda$. 
    \begin{align}
        \frac{\partial \Delta \theta}{\partial \lambda} &= \frac{\partial\theta_{p4p}}{\partial \lambda} - \frac{\partial\theta_{ffs}}{\partial \lambda}\\
        &= \frac{-\alpha c'(\theta)/(1+\lambda)^2}{(1-\alpha)u''(\theta)-\alpha\big[c''(\theta)/(1+\lambda)-R''(\theta)\big]} - \frac{-\alpha c'(\theta)/(1+\lambda)^2}{(1-\alpha)u''(\theta)-\alpha\big[c''(\theta)/(1+\lambda)\big]}.
    \end{align}

    Since $\theta_{p4p}>\theta_{ffs}$, $u(\theta_{p4p}>u(\theta_{ffs}$ (that is, less negative). Therefore, the pay-for-performance denominator is less negative than the fee-for-service denominator. Hence, $\frac{\partial\theta_{p4p}}{\partial \lambda} > \frac{\partial\theta_{ffs}}{\partial \lambda}$. This implies that for higher expertise, the response to pay-for-performance policies will be greater.

    To test this empirically, I limit the sample of hospitals to those who always have some type of clinical executive over the sample period. Then, I estimate the difference in readmission and mortality rates after pay-for-performance among hospitals who do and do not have an internal medicine clinical executive. I lose a significant fraction of the sample with this limitation: there are 60 hospitals that consistently have an internal medicine executive or only a different specialty executive. The summary statistics for these samples are in Appendix \ref{app:sumstats}. I present the results from this comparison in Figure \ref{fig:specialty}.

@@

    \citeauthor{gupta2021impacts} (\citeyear{gupta2021impacts}) finds that a large portion of hospital readmission reductions are due to hospitals being less willing to readmit patients from their own hospital, while they will still readmit patients coming from other hospitals. Given the differential response to the HRRP by clinical and non-clinical hospital leaders, the natural question arises as to whether there is a differential propensity to game the system, as considered in previous work (\cite{gupta2021impacts}). While I do not have patient level data, I can test empirically whether the overall patient population of hospitals changes differentially for clinician or non-clinician leaders.

    I use three outcomes to test this hypothesis. First, the CMS Case Mix Index files provide an aggregate measure of patient complexity for each hospital in each year. A higher index implies the patient population in the hospital is more complex in their treatment requirement. Additionally, I consider the population of patients treated in the hospital with public insurance, either Medicare or Medicaid. These outcomes will indicate if certain types of leadership teams attempt to cost shift relative to others after the penalties or rewards took effect.

    I estimate the same synthetic difference-in-differences equation discussed in Section \ref{sec:clinical}. The results of these estimations are presented in Figure \ref{fig:selection}. All of the treatment effects in these estimations are insignificant and null in magnitude, suggesting that there is no evidence of the overall patient population of hospitals changing differentially based on leadership team. This does not mean that there is no change in specific readmission behavior as found in \citeauthor{gupta2021impacts} (\citeyear{gupta2021impacts}), but a limitation of this data only allows me to look at the aggregate patient population as opposed to specifically readmissions.

    \begin{figure}[ht!]
     \caption{Patient Population as Outcomes}
     \centering
     \begin{subfigure}[b]{0.45\textwidth}
         \centering
         \caption{Case Mix Index}
         \includegraphics[width=\textwidth]{Objects/cmi_md_nomd_synth_graph.pdf}
         \label{fig:cmi}
     \end{subfigure}
     \hfill
     \begin{subfigure}[b]{0.45\textwidth}
         \centering
         \caption{Medicare Percent Discharges}
         \includegraphics[width=\textwidth]{Objects/mcare_md_nomd_synth_graph.pdf}
         \label{fig:mcare}
     \end{subfigure}%
        \begin{subfigure}[b]{0.45\textwidth}
         \centering
         \caption{Medicaid Percent Discharges}
         \includegraphics[width=\textwidth]{Objects/mcaid_md_nomd_synth_graph.pdf}
         \label{fig:mcaid}
     \end{subfigure}
        \label{fig:selection}
    \end{figure}

### 2025-11-10 Slides/JobTalk.tex

% Title Slide
\begin{frame}[plain]
% Background image
\begin{tikzpicture}[remember picture,overlay]
    \node
[
    above=0.5cm,
    align=center,
    fill=lightcyan,
    rounded corners,
    inner xsep=15pt,
    inner ysep=10pt, 
    minimum width=0.9\textwidth,
    text width=0.6\textwidth
] (title) at (current page.center)
{
    \LARGE Does Hospital Leadership Matter?  \\[5pt]
    \large Evidence from Pay-for-Performance
};
% Author 
\node[below=1cm] (author) at (title.south){\large Hanna Glenn};
% Date
\node[below=0.25cm] (date) at (author.south){\today};
\end{tikzpicture}

@@

\begin{frame}{Data Collection Process}
\begin{center}
\begin{tikzpicture}[
    node distance=3.1cm and 1.8cm,
    every node/.style={align=center, font=\normalsize},
    process/.style={rectangle, draw, rounded corners, text centered, minimum width=2.5cm, minimum height=2cm, fill=purple!30},
    output/.style={rectangle, draw, text centered, minimum width=2.7cm, minimum height=2.5cm, fill=purple},
    extract/.style={rectangle, draw, rounded corners, text centered, minimum width=2.5cm, minimum height=2cm, fill=sage!30},
    arrow/.style={thick, ->, >=stealth}
]

% Nodes
\node (step1) [process] {Hospital \\EINs};\pause
\node (step2) [process, right of=step1] {Download \\PDFs};
\draw[arrow] (step1) -- (step2); \pause
\node (step3) [process, right of=step2] {Extract\\text};
\draw[arrow] (step2) -- (step3); 
% New green box below step3
\node (extract) [extract, below of=step3, yshift=0.5cm] {Optical\\Character\\Recognition}; 
\draw[arrow] (extract) -- (step3);\pause
\node (step4) [process, right of=step3] {String\\cleaning\\algorithms};
\draw[arrow] (step3) -- (step4); \pause
\node (output) [output, right of=step4, text=white] {Executive\\Name\\Position\\Title};
\draw[arrow] (step4) -- (output);
\end{tikzpicture}
\end{center}
\end{frame}

@@

    \vspace{2mm}
    \begin{wideitemize}
        \item AHA
        \item Hospital Compare
        \item CMS Case Mix Index
        \item Health Care Cost Report Information System (HCRIS)
    \end{wideitemize}

@@

        \vspace{20mm}

        \hyperlink{int_margin}{\beamergotobutton{intensive margin}}
        \hyperlink{cmi}{\beamergotobutton{due to patient selection?}}
        \hyperlink{specifications}{\beamergotobutton{specifications}}
        \hyperlink{specialty}{\beamergotobutton{specialty}}

@@

@@

@@

    \textbf{Future Work}

    \begin{itemize}
        \item Effects of mandated gender and racial diversity on hospital behavior
        \item Networks of executives over time 
        \item Are leaders correlated with hospital fraudulent behavior?
    \end{itemize}
\end{frame}

@@

    \textcolor{black}{\faEnvelope \hspace{2mm} hanna.glenn@emory.edu}

    \textcolor{black}{\faTwitter \hspace{2mm} @hannalynnglenn}
\end{transitionframe}

>>>>

% Title Slide
\begin{frame}[plain]
% Background image
\begin{tikzpicture}[remember picture,overlay]
    \node
[
    above=0.5cm,
    align=center,
    fill=lightcyan,
    rounded corners,
    inner xsep=15pt,
    inner ysep=10pt, 
    minimum width=0.9\textwidth,
    text width=0.6\textwidth
] (title) at (current page.center)
{
    \LARGE Does Hospital Leadership Matter?  \\[5pt]
    \large Evidence from Pay-for-Performance
};
% Author 
\node[below=1cm] (author) at (title.south){\large Hanna Glenn \\
University of Queensland};
% Date
\node[below=0.25cm] (date) at (author.south){\today};
\end{tikzpicture}

@@

%\begin{frame}{Data Collection Process}
%\begin{center}
%\begin{tikzpicture}[
    %node distance=3.1cm and 1.8cm,
    %every node/.style={align=center, font=\normalsize},
    %process/.style={rectangle, draw, rounded corners, %text centered, minimum width=2.5cm, minimum height=2cm, fill=purple!30},
    %output/.style={rectangle, draw, text centered, minimum width=2.7cm, minimum height=2.5cm, fill=purple},
    %extract/.style={rectangle, draw, rounded corners, text centered, minimum width=2.5cm, minimum height=2cm, fill=sage!30},
    %arrow/.style={thick, ->, >=stealth}
%]

% Nodes
%\node (step1) [process] {Hospital \\EINs};\pause
%\node (step2) [process, right of=step1] {Download \\PDFs};
%\draw[arrow] (step1) -- (step2); \pause
%\node (step3) [process, right of=step2] {Extract\\text};
%\draw[arrow] (step2) -- (step3); 
% New green box below step3
%\node (extract) [extract, below of=step3, yshift=0.5cm] {Optical\\Character\\Recognition}; 
%\draw[arrow] (extract) -- (step3);\pause
%\node (step4) [process, right of=step3] {String\\cleaning\\algorithms};
%\draw[arrow] (step3) -- (step4); \pause
%\node (output) [output, right of=step4, text=white] {Executive\\Name\\Position\\Title};
%\draw[arrow] (step4) -- (output);
%\end{tikzpicture}
%\end{center}
%\end{frame}

@@

    \vspace{2mm}
    \begin{wideitemize}
        \item American Hospital Association
        \item Hospital Compare
        \item CMS Case Mix Index
        \item Health Care Cost Report Information System (HCRIS)
    \end{wideitemize}

@@

@@

\begin{frame}{Intensive Margin}
\large
    \begin{equation*}
    y_{ht} = \beta \left(\mathbb{1}\{\rho \leq \eta\}\text{ x post 2012}\right)_{ht} + \gamma_{h} + \delta_t + \epsilon_{ht}
    \end{equation*}

    \begin{equation*}
    y_{ht} = \beta \left(\mathbb{1}\{\rho > \eta\}\text{ x post 2012}\right)_{ht} + \gamma_{h} + \delta_t + \epsilon_{ht}
    \end{equation*}

    \begin{itemize}
        \item $\rho$: fraction of clinical executives
        \item $\eta$: median fraction of clinically trained executives (.2)
    \end{itemize}
\end{frame}

\begin{frame}{Intensive Margin: Readmission Rates}
    \begin{columns}
    \column{0.5\textwidth}
        \centering
        \includegraphics[width=.9\textwidth]{Objects/cont_belowmedread_md_nomd_synth_graph.pdf}
    \column{0.5\textwidth}
        \centering
        \includegraphics[width=.9\textwidth]{Objects/cont_abovemedread_md_nomd_synth_graph.pdf}
    \end{columns}
\end{frame}

@@

\begin{frame}{Other Mechanisms}
\large
    \begin{wideitemize}
        \item Not driven by changes in patient population
        \item Driven primarily by pneumonia
    \end{wideitemize}
\end{frame}

@@

@@

    \textcolor{black}{\faEnvelope \hspace{2mm} hanna.glenn@uq.edu.au}

\end{transitionframe}
